---
title: 'Metacademy Machine Learning Capstone 0: MPG Prediction'
author: "Joshua Gardner"
date: "May 17, 2016"
output: html_document
---

```{r}
library(plyr)
library(tidyr)
```

#Preprocessing

The task here simply asks developers to predict the car MPG using nine attributes in the original dataset shown below:

    1. mpg:           continuous
    2. cylinders:     multi-valued discrete
    3. displacement:  continuous
    4. horsepower:    continuous
    5. weight:        continuous
    6. acceleration:  continuous
    7. model year:    multi-valued discrete
    8. origin:        multi-valued discrete
    9. car name:      string (unique for each instance)

A selection of the dataset is shown below:

  mpg cylinders displacement horsepower weight acceleration model_year origin                  car_name
1  18         8          307      130.0   3504         12.0         70      1 chevrolet chevelle malibu
2  15         8          350      165.0   3693         11.5         70      1         buick skylark 320
3  18         8          318      150.0   3436         11.0         70      1        plymouth satellite
4  16         8          304      150.0   3433         12.0         70      1             amc rebel sst
5  17         8          302      140.0   3449         10.5         70      1               ford torino
6  15         8          429      198.0   4341         10.0         70      1          ford galaxie 500

To preprocess the data, I read it in, converted cylinders to a factor variable. Additionally, because the car_name variable actually contains two data ponts--the manufacturer and the model--I split that variable into two columns, *make* and *model*, using the **separate()** function from Hadley Wickham's tidyr package.
    
```{r}
mpgdata = read.table("auto-mpg.data", col.names = c("mpg", "cylinders", "displacement", "horsepower", "weight", "acceleration", "model_year", "origin", "car_name"), na.strings = "?")

mpgdata <- mpgdata %>%
    separate(car_name, into = c("make", "model"), sep = " ", extra = "merge") %>%
    mutate(cylinders = as.factor(cylinders), make = as.factor(make), model = as.factor(model))

```

